The benefits of reporting critical effect size values\footnote{The manuscript as been posted as a pre-print and has not yet undergone peer-review.}

### 2025-03-04 paper/paper.Rmd
minor updates

Null hypothesis significance testing (NHST) remains the most prominent approach for statistical inference in science, even though there are widespread concerns about the misuse of hypothesis tests [@lakens2021practical; @chow1988significance; @cortina1997logic; @hagen1997praise; @haig2017tests; @miller2017hypothesis; @nickerson2000null; @cohen1994; @gigerenzer2004null; @krueger2001null]. Over the past decades, numerous proposals have emerged to improve the use of hypothesis testing. These include complementing hypothesis tests with effect sizes and their confidence intervals, preregistering hypotheses before the data is collected, and conducting power analyses to determine the sample size \textit{a priori} based on a smallest effect size of interest [@cumming2012statistical; APA, -@apa2020publication]. Researchers often publish studies with low power for medium to small effect sizes [@szucs2017empirical], with the main justification for the sample size being resource limitations [@lakens2022sample]. However, low power makes it challenging to distinguish signal from noise, and combined with the selective reporting of statistically significant results in the published literature leads to an overestimation of effect sizes [@altoe2020enhancing]. Certain types of research questions can be studied by relying on online data collection, which has made it cheaper and more feasible to collect very large samples. Studies with very large samples are well-powered to detect even small effect sizes, but they also require researchers to carefully consider the possibility that an effect might be statistically significant, but practically insignificant [@kirk1996practical; @anvari2023not]. The risk is that researchers will heuristically or opportunistically accept all effects as potentially important.  For this reason "researchers need to explicitly state the mechanisms that can amplify the importance of an observed effect size and the mechanisms that counteract it" [@anvari2023not, p. 504].

@@

With the present article, we propose that researchers compute and report the "critical effect size value(s)" in their empirical articles. This is not intended to replace other strategies aimed at enhancing the NHST approach to inference. Such strategies, such as the emphasis on estimating effect sizes with confidence intervals [@transue2019apa] or the \textit{a priori} planning for statistical power are valuable in their own right. Instead, our proposal serves as a complementary tool, especially beneficial for facilitating the interpretation of results when statistical power deviates from an optimal level (typically falling below, but occasionally exceeding it). Interestingly, critical effect size values can be retrospectively applied even to already published studies. This possible application facilitates potential reframing of the original interpretations. Serving as a tool for retrospective analysis, critical effect size values may enable a reconsideration of the relevance of previously reported findings.

>>>>

Null hypothesis significance testing (NHST) remains the most prominent approach for statistical inference in science, even though there are widespread concerns about the misuse of hypothesis tests [@lakens2021practical; @chow1988significance; @cortina1997logic; @hagen1997praise; @haig2017tests; @miller2017hypothesis; @nickerson2000null; @cohen1994; @gigerenzer2004null; @krueger2001null]. Over the past decades, numerous proposals have emerged to improve the use of hypothesis testing. These include complementing hypothesis tests with effect sizes and their confidence intervals, preregistering hypotheses before the data is collected, and conducting power analyses to determine the sample size \textit{a priori} based on a smallest effect size of interest [@cumming2012statistical; American Psychological Association [APA], -@apa2020publication]. Researchers often publish studies with low power for medium to small effect sizes [@szucs2017empirical], with the main justification for the sample size being resource limitations [@lakens2022sample]. However, low power makes it challenging to distinguish signal from noise, and combined with the selective reporting of statistically significant results in the published literature leads to an overestimation of effect sizes [@altoe2020enhancing]. Certain types of research questions can be studied by relying on online data collection, which has made it cheaper and more feasible to collect very large samples. Studies with very large samples are well-powered to detect even small effect sizes, but they also require researchers to carefully consider the possibility that an effect might be statistically significant, but practically insignificant [@kirk1996practical; @anvari2023not]. The risk is that researchers will heuristically or opportunistically accept all effects as potentially important.  For this reason "researchers need to explicitly state the mechanisms that can amplify the importance of an observed effect size and the mechanisms that counteract it" [@anvari2023not, p. 504].

@@

With the present article, we propose that researchers compute and report the "critical effect size value(s)" in their empirical articles. This is not intended to replace other strategies aimed at enhancing the NHST approach to inference. Such strategies, such as the emphasis on estimating effect sizes with confidence intervals [APA, -@apa2020publication] or the \textit{a priori} planning for statistical power are valuable in their own right. Instead, our proposal serves as a complementary tool, especially beneficial for facilitating the interpretation of results when statistical power deviates from an optimal level (typically falling below, but occasionally exceeding it). Interestingly, critical effect size values can be retrospectively applied even to already published studies. This possible application facilitates potential reframing of the original interpretations. Serving as a tool for retrospective analysis, critical effect size values may enable a reconsideration of the relevance of previously reported findings.

### 2025-03-04 paper/paper.tex
minor updates

Null hypothesis significance testing (NHST) remains the most prominent approach for statistical inference in science, even though there are widespread concerns about the misuse of hypothesis tests (Chow, 1988; Cohen, 1994; Cortina \& Dunlap, 1997; Gigerenzer et al., 2004; Hagen, 1997; Haig, 2017; Krueger, 2001; Lakens, 2021; Miller, 2017; Nickerson, 2000). Over the past decades, numerous proposals have emerged to improve the use of hypothesis testing. These include complementing hypothesis tests with effect sizes and their confidence intervals, preregistering hypotheses before the data is collected, and conducting power analyses to determine the sample size \textit{a priori} based on a smallest effect size of interest (Cumming et al., 2012; APA, 2020). Researchers often publish studies with low power for medium to small effect sizes (Szucs \& Ioannidis, 2017), with the main justification for the sample size being resource limitations (Lakens, 2022). However, low power makes it challenging to distinguish signal from noise, and combined with the selective reporting of statistically significant results in the published literature leads to an overestimation of effect sizes (Altoè et al., 2020). Certain types of research questions can be studied by relying on online data collection, which has made it cheaper and more feasible to collect very large samples. Studies with very large samples are well-powered to detect even small effect sizes, but they also require researchers to carefully consider the possibility that an effect might be statistically significant, but practically insignificant (Anvari et al., 2023; Kirk, 1996). The risk is that researchers will heuristically or opportunistically accept all effects as potentially important. For this reason ``researchers need to explicitly state the mechanisms that can amplify the importance of an observed effect size and the mechanisms that counteract it'' (Anvari et al., 2023, p. 504).

@@

\begin{table}[H]
\centering
\caption{\label{tab:eqsummary}Summary of the critical effect size equations, see text for details.}
\centering
\begin{tabular}[t]{ll}
\toprule
\textbf{Model} & \textbf{Equation}\\
\midrule
One-sample t-test & $d_c = \frac{t_c \times \mbox{SE}_b}{s}$\\
\cmidrule{1-2}
 & $d_c = \frac{t_c \times \mbox{SE}_b}{s_p}$\\
\cmidrule{2-2}
\multirow{-2}{*}{\raggedright\arraybackslash Paired t-test} & $d_{z_c} = \frac{t_c \times \mbox{SE}_b}{s_D}$\\
\cmidrule{1-2}
Between-sample t-test & $d_c = \frac{t_c \times \mbox{SE}_b}{s_p}$\\
\cmidrule{1-2}
 & $r_c = \frac{t_c}{\sqrt{n - 2 + t^2_c}}$\\
\cmidrule{2-2}
\multirow{-2}{*}{\raggedright\arraybackslash Correlation} & $r_{z_c} = \frac{z_c}{\sqrt{n - 3}}$\\
\cmidrule{1-2}
Meta-analysis/Regression & $\beta_c = t_c \times \mbox{SE}_{\beta}$\\
\bottomrule
\end{tabular}
\end{table}

@@

With the present article, we propose that researchers compute and report the ``critical effect size value(s)'' in their empirical articles. This is not intended to replace other strategies aimed at enhancing the NHST approach to inference. Such strategies, such as the emphasis on estimating effect sizes with confidence intervals (Transue, 2019) or the \textit{a priori} planning for statistical power are valuable in their own right. Instead, our proposal serves as a complementary tool, especially beneficial for facilitating the interpretation of results when statistical power deviates from an optimal level (typically falling below, but occasionally exceeding it). Interestingly, critical effect size values can be retrospectively applied even to already published studies. This possible application facilitates potential reframing of the original interpretations. Serving as a tool for retrospective analysis, critical effect size values may enable a reconsideration of the relevance of previously reported findings.

@@

\leavevmode\vadjust pre{\hypertarget{ref-transue2019apa}{}}%
Transue, B. (2019). \emph{APA style 7th edition}.

>>>>

Null hypothesis significance testing (NHST) remains the most prominent approach for statistical inference in science, even though there are widespread concerns about the misuse of hypothesis tests (Chow, 1988; Cohen, 1994; Cortina \& Dunlap, 1997; Gigerenzer et al., 2004; Hagen, 1997; Haig, 2017; Krueger, 2001; Lakens, 2021; Miller, 2017; Nickerson, 2000). Over the past decades, numerous proposals have emerged to improve the use of hypothesis testing. These include complementing hypothesis tests with effect sizes and their confidence intervals, preregistering hypotheses before the data is collected, and conducting power analyses to determine the sample size \textit{a priori} based on a smallest effect size of interest (Cumming et al., 2012; American Psychological Association {[}APA{]}, 2020). Researchers often publish studies with low power for medium to small effect sizes (Szucs \& Ioannidis, 2017), with the main justification for the sample size being resource limitations (Lakens, 2022). However, low power makes it challenging to distinguish signal from noise, and combined with the selective reporting of statistically significant results in the published literature leads to an overestimation of effect sizes (Altoè et al., 2020). Certain types of research questions can be studied by relying on online data collection, which has made it cheaper and more feasible to collect very large samples. Studies with very large samples are well-powered to detect even small effect sizes, but they also require researchers to carefully consider the possibility that an effect might be statistically significant, but practically insignificant (Anvari et al., 2023; Kirk, 1996). The risk is that researchers will heuristically or opportunistically accept all effects as potentially important. For this reason ``researchers need to explicitly state the mechanisms that can amplify the importance of an observed effect size and the mechanisms that counteract it'' (Anvari et al., 2023, p. 504).

@@

\begin{table}[H]
\centering
\caption{\label{tab:eqsummary}Summary of the critical effect size equations, see text for details. When possible, we also included the equation to calculate the critical effect size without computing the standard error. For the paired t-test, this is feasible when using the standard deviation of differences. For the two-sample t-test, it is possible only under the assumption of homogeneity of variances.}
\centering
\begin{tabular}[t]{ll}
\toprule
\textbf{Model} & \textbf{Equation}\\
\midrule
One-sample t-test & $d_c = \frac{t_c \times \mbox{SE}_b}{s} = t_c\sqrt{\frac{1}{n}}$\\
\cmidrule{1-2}
 & $d_c = \frac{t_c \times \mbox{SE}_b}{s_p}$\\
\cmidrule{2-2}
\multirow{-2}{*}{\raggedright\arraybackslash Paired t-test} & $d_{z_c} = \frac{t_c \times \mbox{SE}_b}{s_D} = t_c\sqrt{\frac{1}{n}}$\\
\cmidrule{1-2}
Between-sample t-test & $d_c = \frac{t_c \times \mbox{SE}_b}{s_p} = t_c\sqrt{\frac{1}{n_1} + \frac{1}{n_2}}$\\
\cmidrule{1-2}
Correlation test & $r_c = \frac{t_c}{\sqrt{n - 2 + t^2_c}}$\\
\cmidrule{1-2}
Correlation & $r_{z_c} = \frac{z_c}{\sqrt{n - 3}}$\\
\cmidrule{1-2}
Meta-analysis/Regression & $\beta_c = t_c \times \mbox{SE}_{\beta}$\\
\bottomrule
\end{tabular}
\end{table}

@@

With the present article, we propose that researchers compute and report the ``critical effect size value(s)'' in their empirical articles. This is not intended to replace other strategies aimed at enhancing the NHST approach to inference. Such strategies, such as the emphasis on estimating effect sizes with confidence intervals (APA, 2020) or the \textit{a priori} planning for statistical power are valuable in their own right. Instead, our proposal serves as a complementary tool, especially beneficial for facilitating the interpretation of results when statistical power deviates from an optimal level (typically falling below, but occasionally exceeding it). Interestingly, critical effect size values can be retrospectively applied even to already published studies. This possible application facilitates potential reframing of the original interpretations. Serving as a tool for retrospective analysis, critical effect size values may enable a reconsideration of the relevance of previously reported findings.

@@
